\classheader{09-20-2017}

\section*{The Cantor set}

The Cantor set is one of those pathological examples in mathematics.  Consider the interval $C_0=[0,1]\subset\mathbb{R}$.  Given $C_k$, define $C_{k+1}$ as $C_k$ with the middle third of each constituent interval removed.  So $$C_1=[0,\frac{1}{3}]\cup[\frac{2}{3},1]$$ $$C_2=[0,\frac{1}{9}]\cup[\frac{2}{9},\frac{1}{3}]\cup[\frac{2}{3},\frac{7}{9}]\cup[\frac{8}{9},1]$$ and so on.

While we don't have a precise definition for what it means to take a limit to $C_\infty$, we can define $C_\infty=\bigcap C_i$ and this is totally fine from a topological point of view.



\definition{The \textbf{Cantor set} is the name given to $C_\infty = \mathcal{C}$.}

The Cantor set is not empty, it's actually uncountable.  To see this, consider all reals in $[0,1]$ expressed in their ternary expansion.  The Cantor set contains all numbers which do not have any $1$s in this representation.  This is obviously an uncountable set.  We also note that the Cantor set doesn't contain any intervals.  

The Cantor set inherits a subspace topology from $\mathbb{R}$.  It is an exercise on the homework to show that the map $f:C_\infty \rightarrow [0,1]$ where we replace all of the $2$s in the ternary representation with $1$s and interpret it as the binary representation of a real number is a continuous function.

The Cantor set has no interior points, so the complement of the Cantor set is dense and open.

\definition{Denote the \textbf{measure} of the set $C_k$ as $mC_k$. Here we'll use measure as `total length', although in a proper, measure-theoretic sense, this definition isn't quite correct.}

What is the measure of the Cantor set?  The measure of $C_0=[0,1]$ is $1$.  We can define a recurrence, where $mC_k=1-\frac{1}{2}\sum\limits_{i=1}^{k}(\frac{2}{3})^i$.  Then, $mC_\infty = 1-\frac{1}{2}\sum\limits_{i=1}^{\infty}(\frac{2}{3})^i = 1-\frac{1}{2}\frac{\frac{2}{3}}{1-\frac{2}{3}}=0$.

What if at each stage we remove a little less than $\frac{1}{3}$ of each interval?  Say we remove $\frac{\alpha}{3}$, where $0<\alpha<1$.  We still get a Cantor-like set, but here we get a recurrence which looks like $mC_k=mC_{k-1}-2^{k-1}\alpha(\frac{1}{3})^k$.  Here, the measure of $C_\infty$ is $1-\alpha$, but its interior is still empty, so its complement is open and dense.

We can even make a Cantor-like set of full measure by putting a smaller copy of the Cantor set into each gap created by removing an interval.  This set is uncountable, has measure $1$, and is nowhere dense.

\definition{Call the function $\frak{c}:\mathcal{C}\rightarrow [0,1]$ such that if $a=0.a_1a_2a_3\dots_t$ is the ternary expansion of an element of the cantor set, $\frak{c}(a)=0.\frac{a_1}{2}\frac{a_2}{2}\frac{a_3}{2}\dots_b$ where we interpret $\frak{c}(a)$ as a binary number, the \textbf{Cantor map}.}

The Cantor map is pretty obviously surjective, but it is not bijective.  To see this, consider $0.0222\dots_t$ and $0.2_t$.  These are not equal (they are representations of $\frac{1}{3}$ and $\frac{2}{3}$, respectively) but the first maps to $0.0111\dots_b$ and the second to $0.1_b$, and both of these are representations of $\frac{1}{2}$, so the map is clearly not injective.  This map is continuous, as we showed in the homework.

We can also think of extending this map to $\tilde{\frak{c}}:[0,1]\rightarrow [0,1]$ where we use the original map on elements of the Cantor set and linear interpolation on points not in the Cantor set.

This leads to a similar map $f:[0,1]\rightarrow [0,1]\times [0,1]$ where if $a=0.a_1a_2a_3a_4\dots_t$, $f(a)=(0,\frac{a_1}{2}\frac{a_3}{2}\dots_b, 0.\frac{a_2}{2}\frac{a_4}{2}\dots_b)$.  This map is also continuous and surjective, but not bijective.  We're now really close to the idea that there could exist a homeomorphism $g:[0,1]\rightarrow [0,1]\times[0,1]$.  In particular, this would mean that dimension is not invariant under homeomorphism and that things like $\mathbb{R}^m$ is homeomorphic to $\mathbb{R}^n$ for all $m,n$.  Can such a homeomorphism exist?  The answer is no, thankfully. 