%
% This is the LaTeX template file for lecture notes for CS294-8,
% Computational Biology for Computer Scientists.  When preparing 
% LaTeX notes for this class, please use this template.
%
% To familiarize yourself with this template, the body contains
% some examples of its use.  Look them over.  Then you can
% run LaTeX on this file.  After you have LaTeXed this file then
% you can look over the result either by printing it out with
% dvips or using xdvi.
%
% This template is based on the template for Prof. Sinclair's CS 270.

\documentclass[twoside]{article}
\usepackage{graphics}
\usepackage{amsfonts}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{IEEEtrantools}
\usepackage{amsmath,amssymb,trimclip,adjustbox}

\mathchardef\mhyphen="2D % Define a "math hyphen"

\usetikzlibrary{chains,fit,shapes}
\setlength{\oddsidemargin}{0.25 in}
\setlength{\evensidemargin}{-0.25 in}
\setlength{\topmargin}{-0.6 in}
\setlength{\textwidth}{6.5 in}
\setlength{\textheight}{8.5 in}
\setlength{\headsep}{0.75 in}
\setlength{\parindent}{0 in}
\setlength{\parskip}{0.1 in}

%
% The following commands set up the lecnum (lecture number)
% counter and make various numbering schemes work relative
% to the lecture number.
%
\newcounter{lecnum}
\renewcommand{\thepage}{\thelecnum-\arabic{page}}
\renewcommand{\thesection}{\thelecnum.\arabic{section}}
\renewcommand{\theequation}{\thelecnum.\arabic{equation}}
\renewcommand{\thefigure}{\thelecnum.\arabic{figure}}
\renewcommand{\thetable}{\thelecnum.\arabic{table}}

%
% The following macro is used to generate the header.
%
\newcommand{\chno}[4]{
   \pagestyle{headings}
   \thispagestyle{plain}
   \newpage
   \setcounter{lecnum}{#1}
   \setcounter{page}{1}
   \noindent
   \begin{center}
   \framebox{
      \vbox{\vspace{2mm}
    \hbox to 6.28in { {\bf CIS 511: Theory of Computation
                        \hfill Apr 18, 2017} }
       \vspace{4mm}
       \hbox to 6.28in { {\Large \hfill Lecture #1: #2  \hfill} }
       \vspace{2mm}
       \hbox to 6.28in { {\it Professor #3 \hfill #4} }
      \vspace{2mm}}
   }
   \end{center}
   \markboth{Lecture #1: #2}{Lecture #1: #2}
   {\bf NB}: {\it These notes are from CIS511 at Penn. The course followed Michael Sipser's \textit{Introduction to the Theory of Computation (3ed)} text.}
   \vspace*{4mm}
}

%
% Convention for citations is authors' initials followed by the year.
% For example, to cite a paper by Leighton and Maggs you would type
% \cite{LM89}, and to cite a paper by Strassen you would type \cite{S69}.
% (To avoid bibliography problems, for now we redefine the \cite command.)
% Also commands that create a suitable format for the reference list.
\renewcommand{\cite}[1]{[#1]}
\def\beginrefs{\begin{list}%
        {[\arabic{equation}]}{\usecounter{equation}
         \setlength{\leftmargin}{2.0truecm}\setlength{\labelsep}{0.4truecm}%
         \setlength{\labelwidth}{1.6truecm}}}
\def\endrefs{\end{list}}
\def\bibentry#1{\item[\hbox{[#1]}]}

%Use this command for a figure; it puts a figure in wherever you want it.
%usage: \fig{NUMBER}{SPACE-IN-INCHES}{CAPTION}
\newcommand{\fig}[3]{
			\vspace{#2}
			\begin{center}
			Figure \thelecnum.#1:~#3
			\end{center}
	}
% Use these for theorems, lemmas, proofs, etc.
\newtheorem{theorem}{Theorem}[lecnum]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newenvironment{proof}{{\bf Proof:}}{\hfill\rule{2mm}{2mm}}

% **** IF YOU WANT TO DEFINE ADDITIONAL MACROS FOR YOURSELF, PUT THEM HERE:

\begin{document}
%FILL IN THE RIGHT INFO.
%\lecture{**LECTURE-NUMBER**}{**DATE**}{**LECTURER**}{**SCRIBE**}
\chno{24}{Circuit Complexity}{Sampath Kannan}{Zach Schutzman}
%\footnotetext{These notes are partially based on those of Nigel Mansell.}

% **** YOUR NOTES GO HERE:

% Some general latex examples and examples making use of the
% macros follow.  
%**** IN GENERAL, BE BRIEF. LONG SCRIBE NOTES, NO MATTER HOW WELL WRITTEN,
%**** ARE NEVER READ BY ANYBODY.

\section*{Parallel Computation}

\definition{The \textbf{Parallel Random Access Model} (PRAM) is a model with synchronous processes with shared memory.  Each process can read one cell, perform a computation, and write one cell at each step, synchronously.}

\definition{The \textbf{Circuit Model} is a model where each gate is considered to be a processor.}

\definition{A \textbf{bounded fan-in (fan-out)} model is one in which we limit the fan-in (fan-out) of any gate to be at most some constant.}

We can convert unbounded fan-in AND and OR gates into bounded ones by replacing it with a tree of sufficient depth.

\claim{Circuits where all gates have fan-out $1$ can be thought of as equivalent to formulas.}

We can see this by observing that fan-out of $1$ makes the circuit a tree.  We can then recursively build the formula by considering left and right subtrees separately.

The formula for a circuit with fan-out greater than $1$ is exponentially big.

When considering computation of circuits, there are two parameters we are interested in: the size (number of gates), and the depth (length of the longest path from input to output).

Circuits have a fixed number of inputs.  How can they recognize arbitrary languages?  

\definition{A \textbf{(non-uniform) circuit family} $C$ is a collection of circuits $C_1,C_2\dots$ where we use circuit $C_i$ to compute things on inputs of length $i$.}

Defined this way, we can define a $C$ which computes the Halting Problem.  This is not ideal.

\definition{A \textbf{uniform circuit family} is a collection of circuits $C$ such that there exists a Turing machine $M$ which runs in log-space which on input $1^n$, outputs the circuit $C_n$.}

Now that we stipulate that it is computable by a log-space transducer, we no longer are able to generate a family of circuits to solve the Halting Problem. 

We'll look at simultaneous size-depth bounds (as a function of $n$) on uniform circuit families.  We want $size(n)$ to be at most polynomial and $depth(n)$ to be polylogarithmic ($log^k(n)$).

\definition{The class $NC^i$, for $i\in\mathbb{N}$ is the set of languages recognized by uniform circuit families of size polynomial in $n$ and depth $log^i(n)$.}

\textbf{Example:} The problem of boolean matrix multiplication is like regular matrix multiplication but the entries are boolean values, we replace elementwise multiplication by AND and addition by OR.  This can be done in $NC^1$, as we can design a circuit with $O(n^3)$ gates and $log(n)$ depth to compute this.

This problem (using adjacency matrices) can tell us whether two vertices in a graph are connected by a path of length $2$.

\textbf{Example:} The transitive closure of a directed graph, $TC(G)$ is a graph $G'$ such that there is an edge from $i$ to $j$ in $G'$ if and only if there is a path from $i$ to $j$ in $G$.  This captures all of the reachability information of $G$.  We can compute this by taking $A$ to be the adjacency matrix of $G$ and computing $(I+A)^n$, where $n$ is the number of vertices in $G$.  We need size $n$ and depth $1$ to do the addition, then $log(n)$ multiplications of size $n^3$ and depth $log(n)$.  In total, we have depth $log^2(n)$ and size $n^3log(n)$.  Therefore, transitive closure is in $NC^2$.

\claim{$NC^1\subseteq L \subseteq NL \subseteq NC^2$.}

\begin{proof}
	$L\subseteq NL$ is trivial, we've just proven that $NL\subseteq NC^2$ because $TC$ allows us to compute $PATH$, so $PATH\in NC^2$.  We also know $NC^1\subseteq NC^2$.  We just need to show that $NC^1\subseteq L$.
	
	An $NC^1$ circuit is a polynomial circuit with depth $log(n)$ and polynomial size which is computed by a log-space transducer.  
	
	Let $A$ be a language in $NC^1$.  We will give an algorithm $M$ in $L$ to compute $A$.  On input $w$ first counts the length of $w$ equal to $n$.  It then uses the transducer which generates the circuit family to construct the $n$th circuit $C_n$, then recursively evaluates the circuit on input $w$, starting with the output.  Since the circuit is log-depth, the recursion can be computed by a machine which only remembers one bit at each level to remember the path and one bit for the previous result, if it exists for a total of $2log(n)$ bits.  Hence $NC^1\subseteq L$.
\end{proof}

We also have $NC\subseteq P$.



\end{document}

